---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# openairegraph - Read and manipulate the OpenAIRE Research Graph Dump

<!-- badges: start -->
[![Travis build status](https://travis-ci.org/njahn82/openairegraph.svg?branch=master)](https://travis-ci.org/njahn82/openairegraph)
<!-- badges: end -->

The goal of openairegraph is to provide parsing functions to consume metadata from the OpenAIRE Research Graph.

Dump: Manghi, Paolo, Atzori, Claudio, Bardi, Alessia, Shirrwagen, Jochen, Dimitropoulos, Harry, La Bruzzo, Sandro, … Summan, Friedrich. (2019). OpenAIRE Research Graph Dump (Version 1.0.0-beta) [Data set]. Zenodo. <http://doi.org/10.5281/zenodo.3516918>

Documentation: <https://api.openaire.eu/graph-dumps.html>


## Installation

You can install the development version of openairegraph from GitHub using the remotes package

``` r
remotes::install_github("openairegraph")
```

## Example

OpenAIRE data are provided as XML files.

```{r}
library(xml2)
library(openairegraph)
dump_eg <- system.file("extdata", "multiple_projects.xml", package = "openairegraph")
my_record <- xml2::read_xml(dump_eg)
```

### Parse publication metadata

```{r}
openairegraph::publications_md(my_record)
```

Author infos

```{r}
openairegraph::publications_md(my_record)$authors
```

Linked persistent identifier (PID) to research publication

```{r}
openairegraph::publications_md(my_record)$pids
```

### Linked projects

```{r}
openairegraph::linked_projects(my_record)
```

### Linked Full-Texts

```{r}
oarg_linked_ftxt(my_record)
```

### Affiliation data

```{r}
oarg_linked_affiliations(my_record)
```


## Meta 

Please note that the 'openairegraph' project is released with a
[Contributor Code of Conduct](CODE_OF_CONDUCT.md).
By contributing to this project, you agree to abide by its terms.
