---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# openairegraph - Read and manipulate the OpenAIRE Research Graph Dump with R

<!-- badges: start -->
[![Travis build status](https://travis-ci.org/subugoe/openairegraph.svg?branch=master)](https://travis-ci.org/subugoe/openairegraph)
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://www.tidyverse.org/lifecycle/#experimental)
[![Codecov test coverage](https://codecov.io/gh/subugoe/openairegraph/branch/master/graph/badge.svg)](https://codecov.io/gh/subugoe/openairegraph?branch=master)
<!-- badges: end -->

## About

This R package provides helpers for splitting, de-compressing and parsing OpenAIRE Research Graph dumps, a big scholarly data dump comprising metadata about various kinds of grant-supported research outpus, as well as the relationships between them. 
The package `openairegraph` targets users who wish to conduct their own analysis using the OpenAIRE Research Graph, but are wary of handling its large data dumps.

More information about OpenAIRE Research Graph, the dumps and the documentation of their structure can be found at 

Manghi, Paolo, Atzori, Claudio, Bardi, Alessia, Schirrwagen, Jochen, Dimitropoulos, Harry, La Bruzzo, Sandro, … Summan, Friedrich. (2019). OpenAIRE Research Graph Dump (Version 1.0.0-beta) Zenodo. <http://doi.org/10.5281/zenodo.3516918>

The data model is more thoroughly explained here:

Manghi, Paolo, Bardi, Alessia, Atzori, Claudio, Baglioni, Miriam, Manola, Natalia, Schirrwagen, Jochen, & Principe, Pedro. (2019, April 17). The OpenAIRE Research Graph Data Model (Version 1.3). Zenodo. <http://doi.org/10.5281/zenodo.2643199>

## Currently implemented methods

So far,  `openairegraph` has been tested to work with the H2020 dump, `h2020_results.gz`. The first set provides helpers to split a large OpenAIRE Research Graph data dump into separate, de-coded XML records that can be stored individually. The other set consists of parsers that convert data from these XML files to a tibble.

For a long-form documentation including a use-case, see:

<https://subugoe.github.io/openairegraph/articles/intro_h2020/oaire_graph_post.html>

## Installation

You can install the development version of openairegraph from GitHub using the remotes package

``` r
remotes::install_github("subugoe/openairegraph")
```

## Quick start

The workflow starts with loading a downloaded OpenAIRE Research Graph dump. After that, the package helps you to de-code and split into several locally stored files. Dedicated parser will obtain data from these files. 

### De-code and split OpenAIRE Research Graph dumps

OpenAIRE Research Graph dumps are json-files that contain a record identifier and a [Base64](https://en.wikipedia.org/wiki/Base64)-encoded text string representing the metadata.

```{r}
library(jsonlite)
library(tibble)
# sample file delivered with this package
dump_file <- system.file("extdata", "h2020_results_short.gz", package = "openairegraph")
# a dump file is in json format
loaded_dump <- jsonlite::stream_in(file(dump_file), verbose = FALSE)
tibble::as_tibble(loaded_dump)
```

`openairegraph::oaire_decode()` decodes these strings and saves them locally. It writes out each XML-formatted record as a zip file to a specified folder.

```r
library(openairegraph)
# writes out each XML-formatted record as a zip file to a specified folder
openairegraph::oaire_decode(loaded_dump, limit = 10, records_path = "data/")
```

These files can be loaded using the `xml2` package.

```{r}
library(xml2)
library(openairegraph)
# sample file delivered with this package
dump_eg <- system.file("extdata", "multiple_projects.xml", 
                       package = "openairegraph")
my_record <- xml2::read_xml(dump_eg)
my_record
```

### XML-Parsers

So far, there are four parsers available to consume the H2020 results set:

- `openairegraph::oarg_publications_md()` retrieves basic publication metadata complemented by author details and access status
- `openairegraph::oarg_linked_projects()` parses grants linked to publications
- `openairegraph::oarg_linked_ftxt()` gives full-text links including access information
- `openairegraph::oarg_linked_affiliations()` parses affiliation data

#### Basic publication metadata

```{r}
openairegraph::oarg_publications_md(my_record)
```

Author infos

```{r}
openairegraph::oarg_publications_md(my_record)$authors
```

Linked persistent identifiers (PID) to a research publication

```{r}
openairegraph::oarg_publications_md(my_record)$pids
```

#### Linked projects

```{r}
openairegraph::oarg_linked_projects(my_record)
```

#### Linked Full-Texts

```{r}
openairegraph::oarg_linked_ftxt(my_record)
```

#### Affiliation data

```{r}
openairegraph::oarg_linked_affiliations(my_record)
```

## Meta 

Please note that the `openairegraph` project is released with a
[Contributor Code of Conduct](CODE_OF_CONDUCT.md).
By contributing to this project, you agree to abide by its terms.

License: MIT

Please use the issue tracker for bug reporting and feature requests.

