---
title: "Read and manipulate the OpenAIRE Research Graph Dump with R"
author: "Najko Jahn"
opengraph:
  twitter:
    card: summary
    creator: "@najkoja"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  out.width = "100%"
)
```

The workflow starts with loading a downloaded OpenAIRE Research Graph dump. 
After that, the package helps you to de-code and split into several locally stored files. 
A dedicated parser will obtain data from these files. 


## De-code and split OpenAIRE Research Graph dumps

OpenAIRE Research Graph dumps are json-files that contain a record identifier and a [Base64](https://en.wikipedia.org/wiki/Base64)-encoded text string representing the metadata.

```{r}
library(jsonlite)
library(tibble)
# sample file delivered with this package
dump_file <- system.file(
  "extdata", "h2020_results_short.gz",
  package = "openairegraph"
)
# a dump file is in json format
loaded_dump <- jsonlite::stream_in(file(dump_file), verbose = FALSE)
tibble::as_tibble(loaded_dump)
```

`openairegraph::oarg_decode()` decodes these strings and saves them locally. 
It writes out each XML-formatted record as a zip file to a specified folder.

```{r, eval=FALSE}
library(openairegraph)
# writes out each XML-formatted record as a zip file to a specified folder
dir.create("data")
oarg_decode(loaded_dump, limit = 10, records_path = "data/")
```

These files can be loaded using the `xml2` package.

```{r}
library(xml2)
# sample file delivered with this package
dump_eg <- system.file(
  "extdata", "multiple_projects.xml", 
  package = "openairegraph"
)
my_record <- xml2::read_xml(dump_eg)
my_record
```


### XML-Parsers

So far, there are four parsers available to consume the H2020 results set:

- `openairegraph::oarg_publications_md()` retrieves basic publication metadata complemented by author details and access status
- `openairegraph::oarg_linked_projects()` parses grants linked to publications
- `openairegraph::oarg_linked_ftxt()` gives full-text links including access information
- `openairegraph::oarg_linked_affiliations()` parses affiliation data


#### Basic publication metadata

```{r}
openairegraph::oarg_publications_md(my_record)
```

Author infos

```{r}
openairegraph::oarg_publications_md(my_record)$authors
```

Linked persistent identifiers (PID) to a research publication

```{r}
openairegraph::oarg_publications_md(my_record)$pids
```


#### Linked projects

```{r}
openairegraph::oarg_linked_projects(my_record)
```


#### Linked Full-Texts

```{r}
openairegraph::oarg_linked_ftxt(my_record)
```


#### Affiliation data

```{r}
openairegraph::oarg_linked_affiliations(my_record)
```
